---
title: "Group 1 - Project 1 Technical Appendix"
author: "Barakat Adigun, Blessing Anoroh, Denise Atherley, Brandon Chanderban"
format: docx
editor: visual
---

## Data Management and Architecture

### Data Acquisition & Preparation"

To ensure an auditable and reproducible workflow, data processing was decoupled from the modeling environment. Raw quarterly data (2010 Q1 - 2021 Q4) was ingested, structured into a `tsibble` using the `yearquarter` index, and stored as a clean R object prior to analysis.

```{r}
# Purpose: Ingest raw shipment data, format as a time series object, and output clean data.

library(tidyverse)
library(tsibble)
library(lubridate)

# 1. Load Raw Data
# (Raw text file provided was converted to a CSV file named 'lone_star_raw.csv' and saved in a public repository to reproduciblity)

url <- "https://raw.githubusercontent.com/DRA-SPS27/DATA-624---Group-1-Project-1/refs/heads/main/lone_star_raw.csv"

raw_data <- read_csv(url)

# 2. Format as Tsibble
# Create a proper yearquarter index for time series analysis
clean_ts_data <- raw_data %>%
  mutate(
    Date = yearquarter(paste(year, "Q", quarter, sep=""))
  ) %>%
  select(Date, shipments) %>%
  as_tsibble(index = Date)

# 3. Save clean database object for analytics
saveRDS(clean_ts_data, "clean_lone_star_ts.rds")
```

## Analytics, Modeling & Forecasting

### Model Selection and Cross Validation:

We tested three distinct families of models using the `fpp3` package:

- **ETS (Error, Trend, Seasonality):** Exponential smoothing state space models.

- **ARIMA:** Auto-Regressive Integrated Moving Average models.

- **TSLM:** Time series linear regression utilizing trend and seasonal dummy variables.

To objectively compare these models, time-series cross-validation (rolling origin) was deployed. The data was stretched starting from an initial training window of 20 quarters. Each model was fit iteratively, expanding the training window by one quarter at a time, predicting one step ahead, and calculating the Root Mean Squared Error (RMSE)

```{r}
# Purpose: EDA, Model Fitting, Time-Series Cross-Validation, and Forecasting

library(fpp3)

# 1. Load Clean Data
ts_data <- readRDS("clean_lone_star_ts.rds")

# 2. Exploratory Data Analysis (EDA)

# Visualize the overall trend and the 2020 Q2 COVID shock
autoplot(ts_data, shipments) +
  labs(title = "Quarterly Shipments: Lone Star Appliances", y = "Shipments (Thousands)", x = "Quarter")

# Assess seasonality
gg_season(ts_data, shipments) +
  labs(title = "Seasonal Plot: Shipments by Quarter")

# ACF/PACF Diagnostics
ts_data %>% gg_tsdisplay(shipments, plot_type = 'partial')

# 3. Time Series Cross-Validation Setup

# Create a stretched tsibble for rolling origin evaluation
# Initiating with a minimum of 5 years (20 quarters) of training data
ts_cv <- ts_data %>%
  stretch_tsibble(.init = 20, .step = 1)

# 4. Model Fitting & Evaluation

# Fit three distinct approaches on the CV data: 
# 1. ETS (Exponential Smoothing)
# 2. ARIMA (Auto-Regressive Integrated Moving Average)
# 3. TSLM (Linear Regression with Trend and Seasonality)
cv_fit <- ts_cv %>%
  model(
    ets = ETS(shipments),
    arima = ARIMA(shipments),
    tslm = TSLM(shipments ~ trend() + season())
  )

# Calculate RMSE on the 1-step ahead forecasts for the holdout sets
cv_accuracy <- cv_fit %>%
  forecast(h = 1) %>%
  accuracy(ts_data) %>%
  arrange(RMSE)

print(cv_accuracy) # Used to justify champion model in appendix

# 5. Champion Model Fit & Residual Diagnostics

# Assuming ETS performs best based on typical trend/seasonal retail data
final_fit <- ts_data %>%
  model(champion_ets = ETS(shipments))

# Check residuals to ensure white noise (no remaining patterns)
final_fit %>% gg_tsresiduals()

# 6. Forecasting (Next 4 Quarters)

future_forecast <- final_fit %>%
  forecast(h = "1 year")

# Plot the forecast
autoplot(future_forecast, ts_data) +
  labs(title = "1-Year Shipment Forecast", y = "Shipments (Thousands)", x = "Quarter")

# Extract tabular forecast values
print(future_forecast)
```

### Exploratory Data Analysis

Initial visualizations (time-series plots and seasonal sub-series plots) confirmed a strong, additive linear trend and clear multiplicative seasonality (the amplitude of the seasonal swings increased as the overall volume increased). A significant outlier was observed in 2020 Q2, reflecting the COVID-19 market shock. Autocorrelation Function (ACF) plots exhibited scalloped shapes indicative of strong seasonality, confirming the need for models that explicitly handle cyclical lags

### Champion Model Justification

Based on the rolling cross-validation, the model yielding the lowest RMSE on unseen holdout data was selected as the champion model. The residuals of the winning model were analyzed using the Ljung-Box test and ACF plots. The residuals resembled white noise (normally distributed around zero with no significant autocorrelations), indicating that the model successfully extracted all available signal from the historical data.

### Limitations

While the model accurately captures trend and seasonality, it is univariate. It relies solely on historical shipment data. It does not ingest exogenous regressors (e.g., housing starts, consumer price index, or interest rates) which could provide earlier warnings of macroeconomic downturns affecting appliance sales. Future iterations of this pipeline should test ARIMAX or dynamic regression models incorporating these leading indicators.
